Prediction of 2024 US election ...

### 2024-11-03 paper/paper.qmd
first verision of data section and made updates to appendix and simulation

library(tidyverse)
library(palmerpenguins)
library(here)
library(tibble)
library(knitr)
library(arrow)
library(rstanarm)
library(brms)
library(dplyr)
library(lubridate)
```

# Introduction {#sec-intro}
Election result forecasting has become an essential tool for analysts in political science and the public to predict the outcome of democratic process, such as the presidential election in the United States. Traditionally, individual polls have been used as a snapshot of voter sentiment, but they only reflect temporary changes in the performance of contestants, instead of a precise estimation of the election result. As discussed by @pasek and @blumenthal, the aggregation of multiple polls, or "poll-of-polls," has become a popular technique to reduce individual survey errors and provide more accurate election forecasts. However, the traditional poll aggregation does not reflect dynamics of an election, especially with differences in quality of polls. A more adaptable model is required to predict the election result based on both polling data and additional variables, such as sample size and pollster credibility.

We build a hybrid election forecasting model following the strategies mentioned by @pasek. As @pasek described in their article, aggregation involves determining which surveys are worth including, as well as selecting, combining and averaging results from multiple polls to reduce individual biases and errors. Prediction modeling adds other data to the model that predicts election outcomes based on current dynamics. Hybrid models like the Bayesian approach incorporates prior beliefs based on historical data or expert knowledge and new evidence like economic updates to dynamically adjust the forecast as the campaign progresses.

In this paper, we predict the 2024 us election result with the hybrid election forcasting model. We incorporate aggregation by filtering the polls on @fivethirtyeight by numeric grade that indicates pollster’s reliability, prediction that incorporates social and economic indicators including unemployment rates and abortion rates, and hybrid approaches that uses Bayesian techniques which combines historical data such as the 2016 election data, allowing for a dynamic prediction of the U.S. presidential election.

@@

The data gathering and analysis is done in R [@citeR] with the following packages: knitr [@knitr], tidyverse [@tidyverse], ggplot2 [@ggplot2], dplyr [@dplyr], arrow [@arrow], here [@here] and lubridate [@lubridate].

@@

## Explore the data

@@

The dataset we obtained from @fivethirtyeight is accumulated from multiple polls and surveys. According to FiveThirtyEight, they aggregate polling data conducted by other firms and organizations that meets their methodological and ethical standards [@morris_trump_2024]. The dataset contains a list of questions in polls and surveys and their results. When a new poll is conducted, the poll is appended to the dataset. FiveThirtyEight assign the poll a poll_id, and each question is assigned a question_id.

In each of the polls recorded by FiveThirtyEight, all options of each question are recorded in candidate_names, while the proportion of respondents choosing that option is recorded in pct as a percentage. The pct is our primary response variable.

There are limitations in these measurements. The differences in sampling methods, wording in survey questions and systematic biases are all reflected in the outcome [@radcliffe_538s_2023]. Thus, in the dataset, FiveThirtyEight includes other variables, such as pollscore and sample_size for their prediction model in addition to the polling results for transparency and accuracy. As an example, pollscore indicates the pollster’s reliability rating, a low pollscore indicates a higher reliability rating [@silver_polls_2008].

These variables combined allow researchers to reliably analyze and predict the 2024 US election result over time.

@@

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Sample of cleaned US election data 
#| echo: false
#| warning: false
cleaned_data = read_parquet(here("data/02-analysis_data/cleaned_data.parquet"))

cleaned_data |>
  select(pct, sample_size, pollscore, days_taken_from_election, state, methodology, candidate_name) |>
  head(6) |>
  kable(
    col.names = c("pct", "sample_size", "pollscore", "days_taken_from_election","state", "methodology", "candidate_name"),
    booktabs = TRUE
    ) 
```

@@

```{r}
#| label: fig-clean-data-histogram-Harris
#| fig-cap: the average PCT vs State for Harris
#| echo: false
#| warning: false
#| height: 40

Harris_data = cleaned_data[cleaned_data$candidate_name == "Kamala Harris", ]
Harris_data |>
  group_by(state) |>
  summarise(avg_pct = mean(pct, na.rm = TRUE)) |>
  ggplot(mapping = aes(x = avg_pct, y = state)) + 
  geom_col()  +
  theme_minimal() +
  labs(
    x = "State", y = "support percentage(pct)", title = "average support percentage(pct) for Harris per state"
  )

# cleaned_data %>%
#   mutate(month = factor(month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
#                                           "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))) %>%
#   ggplot(mapping = aes(x = month, y = total_decedents)) + 
#   geom_bar(stat = "identity") +
#   theme_minimal() +
#   labs(
#     x = "Month", 
#     y = "Total Decedents"
#   )
```

```{r}
#| label: fig-clean-data-histogram-trump
#| fig-cap: the average PCT vs State for Trump
#| echo: false
#| warning: false
#| height: 40

Trump_data = cleaned_data[cleaned_data$candidate_name == "Donald Trump", ]
Trump_data |>
  group_by(state) |>
  summarise(avg_pct = mean(pct, na.rm = TRUE)) |>
  ggplot(mapping = aes(x = avg_pct, y = state)) + 
  geom_col()  +
  theme_minimal() +
  labs(
    x = "State", y = "support percentage(pct)", title = "average support percentage(pct) for Harris per state"
  )

@@

# Model

## Basic Model
```{r}
just_harris_data = Harris_data |> na.omit()
model_MLR = lm(pct ~ pollscore + days_taken_from_election + methodology, data = just_harris_data)
summary(model_MLR)
just_harris_data <- just_harris_data |> mutate(fitted_value = predict(model_MLR),  num_harris = round((pct / 100) * sample_size, 0))

predictions = predict(model_MLR, just_harris_data)

ggplot(just_harris_data, aes(x = end_date)) +
  geom_point(aes(y = pct), color = "black") +
  geom_line(aes(y = fitted_value), color = "blue", linetype = "dotted") +
  facet_wrap(vars(methodology)) +
  theme_classic() +
  labs(y = "Harris percent", x = "Date", title = "Linear Model: pct ~ end_date + pollster")
```

```{r}
baye_model_data = just_harris_data
baye_model_data$pct = as.factor(baye_model_data$pct)
baye_model_data$pollscore = as.factor(baye_model_data$pollscore)
baye_model_data$days_taken_from_election = as.factor(baye_model_data$days_taken_from_election)
baye_model_data$methodology = as.factor(baye_model_data$methodology)
baye_model_data$pct <- as.numeric(baye_model_data$pct)
# Define the Bayesian model with brms

formula <- pct ~ pollscore + days_taken_from_election + (1 | methodology) + (1 | state)

priors = normal(0, 2.5, autoscale = TRUE)

bayesian_model_1 <- stan_glmer(
  formula = formula,
  data = just_harris_data,
  family = gaussian(),
  prior = priors,
  prior_intercept = priors,
  seed = 123,
  cores = 4,
  adapt_delta = 0.95
)
pp_check(bayesian_model_1)
summary(bayesian_model_1)

# Plot random effects
plot(bayesian_model_1, pars = "(Intercept)", prob = 0.95)

# Model

@@

>>>>

library(tidyverse)
library(here)
library(tibble)
library(knitr)
library(maps)
library(mapproj)
light_blue <- "#ADD8E6"
dark_blue <- "#6688CC"
light_red <- "#FF7777"
dark_red <- "#B22222"
purple <- "#9370DB"
```

# Introduction {#sec-intro}
Election result forecasting has become an essential tool for analysts in political science and the public to predict the outcome of democratic process, such as the presidential election in the United States. Traditionally, individual polls have been used as a snapshot of voter sentiment, but they only reflect temporary changes in the performance of contestants, instead of a precise estimation of the election result. As discussed by @pasek and @blumenthal, the aggregation of multiple polls, or "poll-of-polls," has become a popular technique to reduce individual survey errors and provide more accurate election forecasts. However, the traditional poll aggregation does not reflect dynamics of an election, especially with real-time changes and the introduction of new data. This creates a gap for a more adaptable model to predict the election result based on both polling data and additional variables, such as historical data and economic indicators.

This paper fills the gap by building a hybrid election forecasting model following the strategies mentioned by @pasek. As @pasek described in their article, aggregation involves determining which surveys are worth including, as well as selecting, combining and averaging results from multiple polls to reduce individual biases and errors. Prediction modeling adds other data to the model that predicts election outcomes based on current dynamics. Hybrid models like the Bayesian approach incorporates prior beliefs based on historical data or expert knowledge and new evidence like economic updates to dynamically adjust the forecast as the campaign progresses.

In this paper, we aim to predict the 2024 us election result with the hybrid election forcasting model. We incorporate aggregation by filtering the polls on @fivethirtyeight by numeric grade that indicates pollster’s reliability, prediction that incorporates social and economic indicators including unemployment rates and abortion rates, and hybrid approaches that leverages Bayesian techniques which combines historical data such as the 2016 election data, allowing for a dynamic prediction of the U.S. presidential election.

@@

@@

@@

In this dataset, each row represents a polling question that records the variables of interest. Each entry allows us to explore the real-world relationships between polling factors and the support percentage (`pct`) for the candidates Kamala Harris and Donald Trump. This dataset enables an analysis of how various polling characteristics influence the reported support levels for the candidates we are focused.

@@

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Sample of cleaned US election data 
#| echo: false
#| warning: false
cleaned_data = read_csv(here("data/02-analysis_data/cleaned_data.csv"))

cleaned_data |>
  select(pct, sample_size, pollscore, days_taken_from_election, state, methodology, candidate_name) |>
  head(6) |>
  kable(
    col.names = c("pct", "sample_size", "pollscore", "days_taken_from_election","state", "methodology", "Candidate Name"),
    booktabs = TRUE
    ) 
```

### States included in anaylsis

After the data cleaning process, 21 states had no polling data. A table showing the number of polls for each state, including those without any polls, is provided in @tbl-state-polls-count.

This absence of polling data is not a significant concern due to the structure of the United States Electoral College (explained in detail in the Appendix). The states lacking polling data have consistently followed historical voting patterns, so predicting the winning candidate in those states is unnecessary

@@

```{r}
#| label: fig-avg-pct-state
#| fig-cap: Historical State Voting Trends Are Maintined in 2024
#| echo: false
#| warning: false
#| 
cleaned_data <- cleaned_data %>% mutate(candidate_name =
  if_else(!(candidate_name %in% c("Kamala Harris", "Donald Trump")), "Other", candidate_name))

cleaned_data %>%
  filter(candidate_name %in% c("Kamala Harris", "Donald Trump")) %>%
  group_by(state, candidate_name) %>%
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop") %>%
  ggplot(mapping = aes(x = avg_pct, y = state, fill = candidate_name)) + 
  geom_col(position = "stack") +
  theme_minimal() +
  labs(
    x = "Average Polled Support (%)", 
    y = "State",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = light_blue, "Donald Trump" = light_red)) +
  geom_vline(aes(xintercept = 50), color = purple, size = 1) +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  theme(axis.text.y = element_text(size = 8))

```
In figure @fig-avg-pct-state, historically Democratic are polling for Kamala and historically Republican states are polling for Trump. Similarly, historically swing states also appear to be close, for instance Michigan (46.9% Trump, 47.5% Harris), Nevada (47.7% Trump, 47.9% Harris), and Pennsylvania (46.9% Trump, 48.2% Harris) all are close to an even split.

```{r}
#| label: fig-avg-pct-methodology
#| fig-cap: Most Metholodgies Do Not Bias Towards a Candidate
#| echo: false
#| warning: false
#| 
cleaned_data <- cleaned_data %>% mutate(candidate_name =
  if_else(!(candidate_name %in% c("Kamala Harris", "Donald Trump")), "Other", candidate_name))

cleaned_data %>%
  filter(candidate_name %in% c("Kamala Harris", "Donald Trump")) %>%
  filter(!is.na(methodology)) %>% 
  group_by(methodology, candidate_name) %>%
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop") %>%
  ggplot(mapping = aes(x = avg_pct, y = methodology, fill = candidate_name)) + 
  geom_col(position = "stack") +
  theme_minimal() +
  labs(
    x = "Average Polled Support (%)", 
    y = "Methodology",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = light_blue, "Donald Trump" = light_red)) +
  geom_vline(aes(xintercept = 50), color = purple, size = 1) +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  theme(axis.text.y = element_text(size = 8))

```
Figure @fig-avg-pct-methodology indicates minimal bias in polling percentages when comparing different polling methodologies, as most polls show an even distribution of support between candidates. However, polls that incorporate IVR (Interactive Voice Response) and online panels tend to exhibit greater variability, often resulting in more pronounced gaps in support between Donald Trump and Kamala Harris, depending on the specific poll.

## Relationship Between Variables

```{r}
#| label: fig-pollscore-samplesize
#| fig-cap: More Reliable Pollsters Have Larger Sample Sizes in their Polls
#| echo: false
#| warning: false

# removing outliers and na's
graph_data <- cleaned_data %>% 
  filter(!is.na(pollscore) & !is.na(sample_size))

outlier_top <- (quantile(graph_data$sample_size, 0.75) + 
                  IQR(graph_data$sample_size)*1.5)

outlier_bottom <- (quantile(graph_data$sample_size, 0.25) - 
                  IQR(graph_data$sample_size)*1.5)

graph_data <- cleaned_data %>%
  filter(sample_size < outlier_top & sample_size > outlier_bottom)

graph_data %>% filter(!is.na(pollscore) & !is.na(sample_size)) %>% 
  ggplot(aes(x = pollscore, y = sample_size)) +
  geom_point(alpha = 0.33) + geom_smooth(method = "lm", se = FALSE, color = light_red)

# this model is really here just for explain, this realtionship should not
# be treated as lienar
cor(graph_data$sample_size, graph_data$pollscore)
lm_score_sample <- lm(sample_size ~ pollscore, data = graph_data)
summary(lm_score_sample)

```
Figure @fig-pollscore-samplesize illustrates a weak positive correlation between a pollster's pollscore and the sample size of their poll. The figure also shows that most polls have a sample size around 800 to 1200 participants. It is also important to note that a few polls with exceptionally large sample sizes were excluded from the graph due to their status as clear outliers.

```{r}
#| label: fig-pollsamplesize-canidate-pct
#| fig-cap: The Sample Size of a Poll does not impact a Candidate's Voting Percentage
#| echo: false
#| warning: false

cleaned_data %>% 
  filter(!is.na(sample_size) & !is.na(pct)) %>% 
  ggplot(aes(x = sample_size, y = pct)) +
  geom_point(alpha = 0.4) + 
  geom_smooth(method = "lm", se = FALSE, color = light_red) +
  labs(x = "Sample Size of the Poll",
       y = "Candidate Percentage") + facet_wrap(~candidate_name)

```
Based on @fig-pollsamplesize-canidate-pct, there does not seem to be a relationship between the sample size of a poll the percentage of a candidate.

```{r}
#| label: fig-pollscore-canidate-pct
#| fig-cap: More Reliable Pollsters Score Higher For Trump
#| echo: false
#| warning: false

cleaned_data %>% 
  filter(!is.na(pollscore) & !is.na(pct)) %>% 
  ggplot(aes(x = pollscore, y = pct)) +
  geom_point(alpha = 0.4) + 
  geom_smooth(method = "lm", se = FALSE, color = light_red) +
  labs(x = "Pollscore",
       y = "Candidate Percentage") + facet_wrap(~candidate_name)

```
Figure @fig-pollscore-candidate-pct depicts the relationship between a pollster's poll score and a candidate's percentage. For Donald Trump, a negative correlation is observed, indicating that pollsters with lower poll scores tend to assign him a higher percentage of support. Conversely, Kamala Harris shows the opposite trend: as the poll score decreases, her percentage tends to rise. This suggests that more reliable polls, characterized by lower poll scores, report higher support for Trump compared to less reliable pollsters.

```{r}
#| label: tbl-methodology-candidate-support
#| fig-cap: pct vs methodology
#| echo: false
#| warning: false

cleaned_data %>% 
  filter(!is.na(methodology) & !is.na(pct)) %>% 
  group_by(methodology, candidate_name) %>% 
  summarise(methodology, candidate_name, n = n(), avg_pct = mean(pct), sd = sd(pct)) %>% 
  arrange(methodology, candidate_name) %>% unique()

```
Based on @tbl-methodology-candidate-support, we might need to remove methodology from the model.

```{r}
#| label: tbl-state-candidate-support
#| fig-cap: pct vs methodology
#| echo: false
#| warning: false

cleaned_data %>% 
  filter(!is.na(state) & !is.na(pct)) %>% 
  group_by(state, candidate_name) %>% 
  summarise(state, candidate_name, n = n(), avg_pct = mean(pct), sd = sd(pct)) %>% 
  arrange(state, candidate_name) %>% unique()

@@

Based on @tbl-state-candidate-support, state might have to be a removed variable.

```{r}
#| label: fig-polldate-canidate-pct
#| fig-cap: Both Canidate Percentages Are Gaining Support Closer to the Election
#| echo: false
#| warning: false

cleaned_data %>% 
  filter(!is.na(days_taken_from_election) & !is.na(pct)) %>% 
  ggplot(aes(x = days_taken_from_election, y = pct)) +
  geom_point(alpha = 0.4) + 
  geom_smooth(method = "lm", se = FALSE, color = light_red) +
  labs(x = "Timing of Polls Relative to Election Day (Days Prior to the Election)",
       y = "Candidate Percentage") + facet_wrap(~candidate_name)

```
Based on @fig-polldate-canidate-pct, Trump and Harris are getting more votes in polls that are done closer to the election. This is a result of non-major candidate support rapidly decreasing. Specifically, Trump support is increasing at a faster rate than Harris. However, the polls in general show a slight lead for Harris throughout the last 100 days.

The rapid decline in third party support could be due to Robert F. Kennedy dropping out of the race (this sentences should probably be in results or discussion section).

\newpage
# Model

@@

# Overview of American Electon System

# Brief Description of American Federal Goverment

The American federal or national government is split into three branches, executive., legislative, and judical The executive branch includes the president and the military. The legislative branch include two subgroups, the House of Representatives and the Senate. These two subgroups create laws. Every state has two Senators and one Representative per approximately 750,000 people. The judicial branch is court system.

## What is the Electoral College {#sec-electoral-college}

The Electoral College is the system used in the United States to elect the president and vice president. Instead of a direct popular vote, each state is allocated a certain number of electors based on its representation in Congress (the total of its Senators and Representatives). When voters cast their ballots, they are actually voting for a slate of electors pledged to a candidate. The candidate who receives a majority of electoral votes (270 out of 538) wins the presidency. This system means that winning the popular vote in a state generally results in winning all of that state's electoral votes. The only exception are the states of Maine and Nebraska, who award electoral votes by congressional district, with two additional votes given to the statewide winner.

The Electoral College results in some states being unnecessary to campaign in, as their strong historical voting pasterns towards either Democrats or Republicans make them unlikely to change, regardless of campaign efforts. Therefore, for statisticians, polling information from these states may not be that useful when trying to predict the outcome of an election. On the other hand, states that can vote either Democratic or Republican (swing states) are immensely important when predicting an election. As a result, campaigns spend hundreds of millions of dollars campaigning and understanding voters there.

```{r}
#| label: fig-currentstateofelection
#| fig-cap: 2024 U.S. Presidential Election State Forecast Map
#| echo: false
#| warning: false

# Define your colors for each state (replace these colors with your desired ones)
state_colors <- c(
  alabama = dark_red,
  arizona = purple,
  arkansas = dark_red,
  california = dark_blue,
  colorado = dark_blue,
  connecticut = dark_blue,
  delaware = dark_blue,
  `district of columbia` = dark_blue,
  florida = light_blue,
  georgia = purple,
  idaho = dark_red,
  illinois = dark_blue,
  indiana = dark_red,
  iowa = light_red,
  kansas = dark_red,
  kentucky = dark_red,
  louisiana = dark_red,
  maine = "gray",
  maryland = dark_blue,
  massachusetts = dark_blue,
  michigan = purple,
  minnesota = light_blue,
  mississippi = dark_red,
  missouri = dark_red,
  montana = dark_red,
  nebraska = "gray",
  nevada = purple,
  `new hampshire` = light_blue,
  `new jersey` = dark_blue,
  `new mexico` = dark_blue,
  `new york` = dark_blue,
  `north carolina` = purple,
  `north dakota` = dark_red,
  ohio = light_red,
  oklahoma = dark_red,
  oregon = dark_blue,
  pennsylvania = purple,
  `rhode island` = dark_blue,
  `south carolina` = dark_red,
  `south dakota` = dark_red,
  tennessee = dark_red,
  texas = light_red,
  utah = dark_red,
  vermont = dark_blue,
  virginia = light_blue,
  washington = dark_blue,
  `west virginia` = dark_red,
  wisconsin = purple,
  wyoming = dark_red
)

# Load the states map data
states_map <- map_data(map = "state")

plotclr <- c(dark_red, light_blue, purple, gray, light_blue, dark_blue)
legend_description <- c(
  "Likely Republican",
  "Lean Republican",
  "Swing States",
  "Lean Democrat",
  "Likely Democrat",
  "Split by District"
)

ggplot(states_map, aes(x = long, y = lat, group = group, fill = region)) +
  geom_polygon(color = "white", size = 0.2) +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  scale_fill_manual(values = state_colors, 
                    breaks = plotclr,
                    labels = legend_description,
                    name = "Election Status") +  
  guides(fill = guide_legend(override.aes = list(size = 5))) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 12, face = "bold"),
        legend.text = element_text(size = 10),
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank())

```
### Legend

```{r}
#| label: tbl-electionmaplegend
#| tbl-cap: Interpretation of Electoral College Map
#| echo: false
#| warning: false

legend_tibble <- tibble(
  Color = c("Dark Red", "Light Red", "Purple", "Light Blue", "Dark Blue", "Gray"),
  Status = c("Republican", "Lean Republican", "Toss-Up", "Lean Democrat", "Democrat", "Split By District")
)

legend_tibble %>% kable(col.names = c("color", "status"))

```
The state statuses presented in this map are based on evaluations from @cnn, @foxnews, and @msnbc, which are generally agreed upon within the American political community. These organizations assessed historical voting patterns and recent polling data to derive their conclusions.

Notably, Nebraska and Maine are indicated in gray due to their delegates being split by district. In Nebraska, the state overall is projected to lean Republican, with the first and third districts also strongly favoring Republican candidates, while the second district leans Democratic. In Maine, the overall expectation is a Democratic leaning, consistent with its first district, though the second district leans Republican. Furthermore, Alaska and Hawaii are not in the map. Alaska is strongly favoring Republicans while Hawaii strongly favors democrats.

The seven generally agreed upon swing states are Pennsylvania, North Carolina, Georgia, Arizona, Nevada, Wisconsin, and Michigan with Texas, Florida, Nebraska District Two, Maine District 2, and Minnesota as the next closest races.

# States Poll Count

```{r}
#| label: tbl-state-polls-count
#| tbl-cap: Polls included in Analysis Per State
#| echo: false
#| warning: false

states <- c(state.name, "National")
polls_per_state_tbl <- tibble(state = states)

polls_per_state <- cleaned_data %>% group_by(state) %>% 
  summarise(polls_count = n_distinct(poll_id))

polls_per_state_tbl <- polls_per_state_tbl %>% 
  left_join(polls_per_state, by = "state") %>% 
  mutate(polls_count = if_else(!is.na(polls_count), polls_count, 0)) %>% 
  select(state, polls_count)

polls_per_state_tbl %>% kable(col.names = c("State", "Number of Polls In Analysis"))

num_states_missing <- count(polls_per_state_tbl %>% 
                              filter(polls_count == 0) %>% 
                              select(state))
